How to publish in Biological Sciences: a guide for the uninitiated

### 2021-12-15 4.05-paywall.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

Being the editor of a society journal means being elected by members of that society, and being responsible to an editorial board, normally made up of the society’s members. Until very recently, and I’m thinking back to my first interactions with editors for my first few publications, submitting to the journal meant producing three (or sometimes more) double spaced hard copies of a manuscript and mailing a large and heavy envelope to the editor. Editors of bigger journals had secretaries dedicated to handling the administration of the paper. Following a telephonic enquiry, the copies were sent out to referees by post and sent back to the editorial office with a typed report often together with the marked up manuscript. Once the editor had received all reports, they communicated their decision back to the corresponding author (i.e. the one to whom correspondence was addressed) and, once accepted, the article went into production. Prior to personal computers being commonplace (only 25 years ago), each journal would have had to have had a publisher to set the type and print the pages. Clearly, this was beyond the scope of individual societies and the publisher was a necessity. Libraries had to pay for copies of journals, as the cost of publishing had to be offset by the society.

@@

After the first issue, I realised that I could not do all of this work indefinitely. I knew that there were publishers who were interested in acquiring the journal into their stable, and I contacted them and started negotiations. In 2011, the first copy of _African Journal of Herpetology_ from a professional publisher emerged, and allowed me to go back to editing the content through peer review.

At the time, I was aware of Open Access and considered this as an option for the journal. Open Access would have required that someone pay for the type setting, and the society would still have to pay for an online dissemination platform, given that they did not have their own platform. This would have meant that authors paid for getting their manuscripts into print. And then, like today, the decision was that our authors would not be willing to pay. Other societies did opt to follow this Open Access route, with the costs being covered by the authors. For some this became an incredibly successful model, with submissions increasing as well as the [Impact Factor](#impactfactor). They demonstrate that Open Access is possible on an independent platform.

@@

In the original publishing model, scholarly societies, such as The Royal Society, paid publishers to type set and print their publications (see Figure \@ref(fig:PhilTransRoySoc)). These in turn were bought by the society for their members and any libraries who subscribed to the society. As institutions and their libraries grew internationally, so too did the subscription base such that there was a profit to be made from the cost of the subscriptions for the paper content. Even back in these early Victorian days of scholarly publishing, publishers to started their own journals (aside from any society) as they recognised their superiority in distribution [@brock1980development]. This model continued over many decades with some publishing houses growing as they acquired more journal titles from societies because there was a lot of profit to be made. The peculiarity of scholarly publishing includes the unpaid nature of authors, editors and reviewers (all of the content), and the non-equivalence of the output: that is to say that just because you have access to articles in _Nature_, they are not equivalent to those being published in _Science_ - as a scientist you need to read articles from both of these publications.

Interestingly, the dawn of the internet was predicted as the downfall of the academic publisher Elsevier, because it promised the ability for academics to share their content for free [@lariviere2015big]. Despite this Elsevier's profit in 2012-13 was USD 2 billion, with a profit margin of ~40% [@lariviere2015big]. This profit margin is so large that some have compared it favourably with street gangs selling drugs [@buranyi2017staggeringly]. In the Biological Sciences, Elsevier, Springer-Nature, Wiley and Taylor-Francis control around 50% of all papers published [@lariviere2015big]. Selling to university libraries is no longer made on print copies, but on packages of electronic journals from each of these big publishers. The deals made by individual libraries were (and are) made in strict confidence, the exact sums are unknown, but the deals made fall into the hundred and thousands of USD for a set of publications that cover (for example) the science content of one of these publishing houses. The penalty for not subscribing is that your researchers reach a [paywall](#paywall) when they try to access a particular paper. You will find, as I have, the luxury of visiting a very wealthy institution and having access to just about every journal title you can think of. If you are not at one of these elite institutions, then you will face a demand for money when you try to access a title that your institution does not subscribe to.

The cost of accessing a single article comes at between 25 and 50 USD [@hagve2020money]. For me, and a great deal of other researchers, there are other ways to [get around the paywall](http://www.howtowriteaphd.org/paywall.html). But for some people, especially members of the public who simply want to access the information, they may simply pay the price. Imagine then the potential profit to be made from a society that has a large back-catalogue of content. There has been a scramble to sign up editors from society journals with tens of thousands of old journal articles at their command. The publisher offers to scan the entire back-catalogue and then offer it on sale. The society gets a paltry 10%, while the publisher fishes with tens of millions of articles. Even if they only sell 10% of these in 1 year, that makes their profit in the millions of USD.

@@

[Open Access](#openaccess) titles came about in the early 2000s as the need for hard printed copies fell away and academics became comfortable handling all of their publications as pdfs. In the Biological Sciences, _PLOS ONE_ was the title that revolutionised the market, quickly becoming the single biggest journal. What caught the eyes of the publishers was the article processing charge (APC) that academics seemed to be prepared to pay. When they first started, _PLOS ONE_ cost ~1300 USD for its APC [@khoo2019article]. Publishers quickly started titles with similar [no impact factor](#noif) goals. Competitive titles began at the same cost as _PLOS ONE_, but once these titles became established, with higher Impact Factors, they quickly increased their APC [@khoo2019article]. Commercial publications (as opposed to non-profit journals) have become adept at pricing their titles toward what libraries and authors are prepared to pay, such that commercial publications are regularly more expensive than non-profit journals [@bjork2015article]. In their analyses of drivers of APC for OA journals, Budzinski et al. [-@budzinski2020drivers] found several variables that explained the vast difference in APCs of different journals. Impact Factor played an important role, but so too did market power of the publishers (presumably the ability to distribute the title and content), the hybrid model ([Gold OA](#gold) or [Hybrid OA](#hybrid)), and the concentration of disciplines that they cover.

Librarians analysed content that was [Open Access](#openaccess) finding that it was more highly cited, and so academics began to look for the potential to publish [Open Access](#openaccess). The publishers were happy to oblige, and in the early days conducted 'double dipping' charging both the APC for the authors, and collecting full revenue for subscription to hybrid journals by universities [@pinfield2016total]. It was only in May 2018 that [Knowledge Unlatched](https://knowledgeunlatched.org/2018/05/knowledge-unlatched-announces-its-fifth-pledging-round-ku-select-2018-2/) formed an alliance to prevent double-dipping in Open Access publications.

The end to double-dipping, as mandated by governments [@pinfield2016total], has led to even more complicated publishing deals known as **transformative publisher agreements** (Table \@ref(tab:profit). With these agreements, researchers from universities that have full subscriptions to certain publishers will receive free Gold OA for any journals in the package. In effect, the price of the OA is offset against the cost of the subscription. As you may have noticed, this will quickly start a bias in the publishing trends of academics from some (middle income) institutions publishing with certain titles, because they receive the benefit of OA. Those from privileged institutions will continue to publish anywhere they can, and the real losers will be those from developing, especially middle-income countries, who cannot afford transformative agreements, and cannot afford to provide their researchers with costs to cover APCs [@measey2018europe]. For these researchers, there will be a reduced set of journals that they can subscribe to. This is the current reality that many of us find ourselves in.

On the face of it, Plan S makes total sense [@schiltz2018science]. If all papers published by European researchers are Open Access, then everyone can read them and there is no more need for paywalls. Unfortunately, the way in which Plan S has been implemented has only been for wealthy European institutions. Others are left to continue to pay a hefty APC price for OA publishing, and their libraries pay a subscription to access journal content [@hagve2020money].

@@

 | Funding source | From society journals | From their own journals | From [Open Access](openaccess.html) | From [copyright](#copyright) |
| ----- | ----- | ----- | ----- | ----- | 
| Old subscription model (mostly obsolete) | X | X | - | - |
| Selling bundles to university libraries (undisclosed sums - millions of USD) | X | X | - | - |
| Selling articles of large back catalogues (~10% goes to society) | X | X | - | - |
| Article Process Charging | - | - | X | - |
| Double dipping  | X | X | - | - |
| Transformative publisher agreements | X | X | X | - |
| Copyright content  | X | X | - | X |

### How could such a profitable business lead to making a loss?
To be fair, we should also consider that some publishers have not (always) been able to turn a profit, even when their turnover is in millions of USD. One such beleaguered publisher is PLOS. Despite growing exponentially in the early noughties, to become the world's biggest academic publisher in 2013, PLOS has not managed to turn a profit since 2015 [@davies2019plos]. At the time of writing (May 2021), the APC on a _PLOS ONE_ article was USD 1749, while other PLOS titles ranged from USD 2100 to 4000 (see [PLOS fees site](https://plos.org/publish/fees/)). This begs the question, how much does it cost to process an article? Data from the PLOS Financial Overview suggests that processing your article costs USD ~315, with another USD ~262 to keep it online.

So where does the rest of the money go? PLOS spend the bulk of your APCs on the editorial aspects of publishing. Remember that PLOS don't pay academic editors, but they do have huge financial burdens of the salaries of their own editorial staff. PLOS is based in San Francisco, California, which just happens to be one of the most expensive cities in the world to live in. Chief among their many options are to raise the costs of your APCs in order to pay their staff, or move their operation to somewhere cheaper. Interestingly, the other two large OA outfits, MDPI and Frontiers, are both based in Switzerland, possibly one of the most expensive countries in the world in which to pay salaried staff.

Most recently, PLOS announced a new funding model to keep themselves afloat that spreads the cost of their APC among the authors, and aims to cap profits at 10% [@else2021open-access]. The new scheme, Community Action Publishing, sees a return to the institutional subscription model where institutions will be charged proportionately at the rate their academics publish in PLOS. Great if your institutions can afford to opt in. If not, you can expect to face your proportion of the APC now as a co-author (instead of only the corresponding author paying).

@@

## What will it take to break the vicious cycle?
We need new models for publishing. Society journals are still kings in this game and ultimately hold the cards for moving away from filling the pockets of publishing companies. What we have seen in recent years is that journals can come from nowhere to become dominant players in the system. Think [_PLOS ONE_](#PLOS), and the even more recent meteoric rise of _Scientific Reports_. These mega Open Access journals didn’t exist ten years ago. And they don’t need to exist ten years from now. What is needed is for the actual costs of publishing (not that currently inflated by publishers) to be covered by the institutions that employ the academics. This could cover type-setting (fulfilling our irrational desire for '[fancy layouts](#layout)' and the additional IT infrastructure (on-line submission system and online dissemination platform). Most (if not all) societies are not-for-profit organisations, and only need to cover the costs of publishing.

@@

If current trends continue, scientists from low-income countries will be granted full fee waivers. Many journals use the [Hinari Eligibility list of countries](http://www.who.int/hinari/eligibility/en/) to separate Group A (free access) and Group B (low cost access – normally billed at a 50% reduction in fees). The lists are made up from five global economic and development criteria. Middle-income countries are missing from these lists, and receive no support for fee waivers. Their governments provide scientists with no means of paying fees. Scientists who pay fees often do so from their own research budgets. The increasing number of journals that charge unjustifiably high publishing costs are forcing middle-income scientists away from Open Access journals.

>>>>

Being the [editor](#gatekeepers) of a society journal means being elected by members of that society, and being responsible to an editorial board, normally made up of the society’s members. Until very recently, and I’m thinking back to my first interactions with editors for my first few publications, submitting to the journal meant producing three (or sometimes more) double spaced hard copies of a manuscript and mailing a large and heavy envelope to the editor. Editors of bigger journals had secretaries dedicated to handling the administration of the paper. Following a telephonic enquiry, the copies were sent out to referees by post and sent back to the editorial office with a typed report often together with the marked up manuscript. Once the editor had received all reports, they communicated their decision back to the corresponding author (i.e. the one to whom correspondence was addressed) and, once accepted, the article went into production. Prior to personal computers being commonplace (only 25 years ago), each journal would have had to have had a publisher to set the type and print the pages. Clearly, this was beyond the scope of individual societies and the publisher was a necessity. Libraries had to pay for copies of journals, as the cost of publishing had to be offset by the society.

@@

After the first issue, I realised that I could not do all of this work indefinitely. I knew that there were publishers who were interested in acquiring the journal into their stable [see @lariviere2015oligopoly], and I contacted them and started negotiations. In 2011, the first copy of _African Journal of Herpetology_ from a professional publisher emerged, and allowed me to go back to editing the content through peer review.

At the time, I was aware of [Open Access](#openaccess1) and considered this as an option for the journal. Open Access would have required that someone pay for the type setting, and the society would still have to pay for an online dissemination platform, given that they did not have their own platform. This would have meant that authors paid for getting their manuscripts into print. And then, like today, the decision was that our authors would not be willing to pay. (Note that today, there are a lot more options for [flipping journals](https://blog.scholasticahq.com/post/how-to-start-flip-open-access-academic-journal).) Other societies did opt to follow this [Open Access](#openaccess1) route, with the costs being covered by the authors. For some this became an incredibly successful model, with submissions increasing as well as the [Impact Factor](#impactfactor). They demonstrate that Open Access is possible on an independent platform.

@@

In the original publishing model, scholarly societies, such as The Royal Society, paid publishers to type set and print their publications (see Figure \@ref(fig:PhilTransRoySoc)). These in turn were bought by the society for their members and any libraries who subscribed to the society's content. As institutions and their libraries grew internationally, so too did the subscription base such that there was a profit to be made from the cost of the subscriptions for the paper content. Even back in those early Victorian days of scholarly publishing, publishers to started their own journals (aside from any society) as they recognised their superiority in distribution [@brock1980development]. This model continued over many decades with some publishing houses growing as they acquired more journal titles from societies because there was a lot of profit to be made. The peculiarity of scholarly publishing includes the unpaid nature of authors, editors and reviewers (all of the content), and the non-equivalence of the output: that is to say that just because you have access to articles in _Nature_, they are not equivalent to those being published in _Science_ - as a scientist you need to read articles from both of these publications.

Interestingly, the dawn of the internet was predicted as the downfall of the academic publisher Elsevier, because it promised the ability for academics to share their content for free [@lariviere2015big]. Despite this, Elsevier's profit in 2012-13 was USD 2 billion, with a profit margin of ~40% [@lariviere2015big]. This profit margin is so large that some have compared it favourably with street gangs selling drugs [@vannoorden2013open; @buranyi2017staggeringly; @logan2017we]. In the Biological Sciences, Elsevier, Springer-Nature, Wiley and Taylor-Francis control around 50% of all papers published [@lariviere2015big]. Selling to university libraries is no longer made on print copies, but on packages of electronic journals from each of these big publishers. The deals made by individual libraries were (and are) made in strict confidence, the exact sums are unknown, but the deals made fall into the hundreds of thousands of USD for a set of publications that cover (for example) the science content of one of these publishing houses. In 2016, it has been estimated that UK universities paid GBP ~200 million to Elsevier alone [see @logan2017we], and global profit from science publishing was estimated at USD 9.4 billion in 2011 [@vannoorden2013open], USD 1.1 billion for Elsevier alone [@fuchs2013diamond]. The penalty for not subscribing is that your researchers reach a [paywall](#paywall) when they try to access a particular paper. You will find, as I have, the luxury of visiting a very wealthy institution and having access to just about every journal title you can think of. If you are not at one of these elite institutions, then you will face a demand for money when you try to access a title that your institution does not subscribe to.

The cost of accessing a single article comes at between 25 and 50 USD [@hagve2020money]. For me, and a great deal of other researchers, there are other ways to [get around the paywall](http://www.howtowriteaphd.org/paywall.html). But for some people, especially members of the public who simply want to access the information, they may simply pay the price. When these are the same tax payers that have paid for the work, it is easy to see why this [paywall](#paywall) has been branded unethical [e.g. @logan2017we]. Imagine then the potential profit to be made from a society that has a large back-catalogue of content. There has been a scramble to sign up editors from society journals with tens of thousands of old journal articles at their command [@lariviere2015oligopoly]. The publisher offers to scan the entire back-catalogue and then offer it on sale. The society gets a paltry 10%, while the publisher fishes with tens of millions of articles. Even if they only sell 10% of these in 1 year, that makes their profit in the millions of USD.

@@

[Open Access](#openaccess) titles came about in the early 2000s as the need for hard printed copies fell away and academics became comfortable handling all of their publications as pdfs. In the Biological Sciences, _PLOS ONE_ was the title that revolutionised the market, quickly becoming the single biggest journal. What caught the eyes of the publishers was the Article Processing Charge ([APC](#APC)) that academics seemed to be prepared to pay. When they first started, _PLOS ONE_ cost ~1300 USD for its [APC](#APC) [@khoo2019article]. Publishers quickly started titles with similar [no impact factor](#noif) goals. Competitive titles began at the same cost as _PLOS ONE_, but once these titles became established, with higher Impact Factors, they quickly increased their [APC](#APC) [@khoo2019article]. Commercial publications (as opposed to non-profit journals) have become adept at pricing their titles toward what libraries and authors are prepared to pay, such that commercial publications are regularly more expensive than non-profit journals [@bjork2015article]. In their analyses of drivers of [APC](#APC) for OA journals, Budzinski et al. [-@budzinski2020drivers] found several variables that explained the vast difference in [APC](#APC)s of different journals. Impact Factor played an important role, but so too did market power of the publishers (presumably the ability to distribute the title and content), the hybrid model ([Gold OA](#gold) or [Hybrid OA](#hybrid)), and the concentration of disciplines that they cover.

Librarians analysed content that was [Open Access](#openaccess) (OA) finding that it was more highly cited, and so academics began to look for the potential to publish OA. The publishers were happy to oblige, and in the early days conducted 'double dipping' charging both the [APC](#APC) for the authors, and collecting full revenue for subscription to hybrid journals by universities [@pinfield2016total]. It was only in May 2018 that [Knowledge Unlatched](https://knowledgeunlatched.org/2018/05/knowledge-unlatched-announces-its-fifth-pledging-round-ku-select-2018-2/) formed an alliance to prevent double-dipping in OA publications.

The end to double-dipping, as mandated by governments [@pinfield2016total], has led to even more complicated publishing deals known as **transformative publisher agreements** (Table \@ref(tab:profit). Unsurprisingly, these deals are said to lead to even greater profits for the publishers [@poynder2020information]. With these agreements, researchers from universities that have full subscriptions to certain publishers will receive free [Gold OA](#gold) for any journals in the package and their authors will be able to publish OA inside the same publisher's journals. In effect, the price of the [APC](#APC) for OA is offset against the cost of the subscription. As you may have noticed, this will quickly start a bias in the publishing trends of academics from some (middle income) institutions publishing with certain titles, because they receive the benefit of OA. Those from privileged institutions will continue to publish anywhere they can, and the real losers will be those from poorer institutions and especially middle-income countries, who cannot afford transformative agreements, and cannot afford to provide their researchers with costs to cover [APC](#APC)s [@measey2018europe]. For these researchers, there will be a reduced set of journals that they can subscribe to. This is the current reality that increasing numbers of researchers find themselves in.

On the face of it, Plan S makes total sense [@schiltz2018science]. If all papers published by European researchers are Open Access, then everyone can read them and there is no more need for paywalls. Unfortunately, the way in which Plan S has been implemented has only been for wealthy European institutions. Others are left to continue to pay a hefty [APC](#APC) price for OA publishing, and their libraries pay a subscription to access journal content [@hagve2020money]. There is a real possibility that the USA and China will take other routes to OA, causing a schism in publishing, more complex agreements and even more profits for publishers [@poynder2020information]. In reality, publishers are using some of the money generated to lobby governments the world over to make sure that they can continue to profit from academia. Plan S is just the sort of result that comes from publishers' effective lobbying.

@@

 | Funding source | From society journals | From their own journals | From [Open Access](openaccess.html) | From [copyright](#copyright) |
| --------- | -- | -- | -- | -- | 
| Old subscription model (mostly obsolete) | X | X | - | - |
| Selling bundles to university libraries (undisclosed sums - millions of USD) | X | X | - | - |
| Selling articles of large back catalogues (~10% goes to society) | X | X | - | - |
| Article Process Charging | - | - | X | - |
| Double dipping  | X | X | - | - |
| Transformative publisher agreements | X | X | X | - |
| Copyright content  | X | X | - | X |

In the last 10 years, big publishers have shifted their strategy from making money on publishing towards collecting and selling data [@posada2018inequality; @brembs2021replacing; @pooley2021surveillance]. Their strategy appears to be buying up the tools that academics use in a work-flow including: discovery, analysis, writing, publication, outreach and assessment. Once this entire work-flow is captured, the publishers can track individual users across it and sell this information to the institutions that employ them [@posada2018inequality; @christl2021digitale], or third parties that may profit from knowing what you read and write [@pooley2021surveillance]. All of this is reminiscent of the Cambridge Analytica scandal which is alleged to have manipulated election results in the UK and USA in 2016. If we want [academia to be captured](#capture), we can continue to keep our eyes on the highest [Impact Factors](#impactfactor), while the big publishers steal our freedom.

### How could such a profitable business lead to making a loss? {#loss}
To be fair, we should also consider that some publishers have not (always) been able to turn a profit, even when their turnover is in millions of USD. One such beleaguered publisher is PLOS. Despite growing exponentially in the early noughties, to become the world's biggest academic publisher in 2013, PLOS has not managed to turn a profit since 2015 [@davies2019plos]. At the time of writing (May 2021), the [APC](#APC) on a _PLOS ONE_ article was USD 1749, while other PLOS titles ranged from USD 2100 to 4000 (see [PLOS fees site](https://plos.org/publish/fees/)). This begs the question, how much does it cost to process an article? Data from the PLOS Financial Overview suggests that processing your article costs USD ~315, with another USD ~262 to keep it online.

So where does the rest of the money go? PLOS spend the bulk of your [APC](#APC)s on the editorial aspects of publishing. Remember that PLOS don't pay academic editors, but they do have huge financial burdens of the salaries of their own editorial staff. PLOS is based in San Francisco, California, which just happens to be one of the most expensive cities in the world to live in. Chief among their many options are to raise the costs of your [APC](#APC)s in order to pay their staff, or move their operation to somewhere cheaper. Interestingly, the other two large OA outfits, MDPI and Frontiers, are both based in Switzerland, one of the most expensive countries in the world in which to pay salaried staff.

Most recently, PLOS announced a new funding model to keep themselves afloat that spreads the cost of their [APC](#APC) among the authors, and aims to cap profits at 10% [@else2021open-access]. The new scheme, Community Action Publishing, sees a return to the institutional subscription model where institutions will be charged proportionately at the rate their academics publish in PLOS. Great if your institutions can afford to opt in. If not, you can expect to face your proportion of the [APC](#APC) now as a co-author (instead of only the corresponding author paying).

@@

## What will it take to break the vicious cycle?
We need new models for publishing. Society journals are still kings in this game and ultimately hold the cards for moving away from filling the pockets of publishing companies. What we have seen in recent years is that journals can come from nowhere to become dominant players in the system. Think [_PLOS ONE_](#PLOS), and the even more recent meteoric rise of _Scientific Reports_. These mega Open Access journals didn’t exist ten years ago. And they don’t need to exist ten years from now. What is needed is for the actual costs of publishing (not that currently inflated by publishers) to be covered by the institutions that employ the academics. This could cover type-setting (fulfilling our irrational desire for '[fancy layouts](#layout)' and the additional IT infrastructure (on-line submission system and online dissemination platform). Most (if not all) academic societies are not-for-profit organisations, and only need to cover the costs of publishing.

@@

If current trends continue, scientists from low-income countries will be granted full fee waivers. Many journals use the [Hinari Eligibility list of countries](http://www.who.int/hinari/eligibility/en/) to separate Group A (free access) and Group B (low cost access – normally billed at a 50% reduction in fees). The lists are made up from five global economic and development criteria. Middle-income countries are missing from these lists, and receive no support for fee waivers. Their governments provide scientists with no means of paying fees. Scientists who pay fees often do so from their own research budgets. The increasing number of journals that charge unjustifiably high publishing costs [@grossmann2021current] are forcing middle-income scientists away from Open Access journals.
